Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-07-06 Manuscript.qmd
added linenumbers

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
---

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------|------------|-------|--------|------|----------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O                         |
|                |                    |                  | Westray (UK)    | QG1-7189-2 | and $\delta$^13^C                            |
|                |                    |                  |                 | QG2-1061-1 | analysis                      |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |

>>>>

---
title: "Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz i Godino
    affiliations:
      - name: Department of Evolutionary Biology, Ecology and Environmental Science, Biodiversity Research Institute (IRBio), University of Barcelona
        address: Avinguda Diagonal, 643
        city: Barcelona
        country: Spain
        postal-code: 08028 
abstract: |
  This study provides a reassessment of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures (SST). While *Patella depressa* along the Spanish shoreline has repeatedly produced reliable correlations between SST and Mg/Ca ratios, this relationship is not the case for other patelloid species. *Patella vulgata*, *Nacella deaurata*, and *N. magellanica* have been studied using Mg/Ca ratios with mixed or contrary results. In this study, we present elemental maps of these three species together with oxygen isotope ratios ($\delta$^18^O) that confirm a good relationship with SST. Our dataset also reassesses a specimen which was previously unsuccessful in providing significant correlations between $\delta$^18^O values and Mg/Ca ratios. By reassessing these species and including modern and archaeological specimens (n=12) from three patelloid species (*P. vulgata*, *N. deaurata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change. As a result, these species can in the future serve to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O and clumped isotope analysis.
keywords: 
  - Sclerochronology
  - Patella
  - Nacella
  - LIBS
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: FALSE
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm
header-includes:
  - \usepackage{lineno}
  - \linenumbers
---

@@

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses                     |
|------|------------|-------|--------|------|----------|
| Modern         | [@Nicastro2020-ih] | *N. deaurata*   | Cambaceres Bay  | ND-1016-3  | $\delta$^18^O and $\delta$^13^C        |
|                |                    |                  | \- Tierra del   | ND-1016-4  | analysis                              |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  |                                       |
|                |                    |                  |                 | NM-1016-3  |                                       |
|                |                    |                  |                 |            |                                       |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    |            | analysis;                             |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.                       |
|                |                    |                  |                 |            |                                       |
|                |                    |                  |                 |            |                                       |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | $\delta$^18^O and $\delta$^13^C       |
|                |                    |                  | Westray (UK)    | QG1-7189-2 |       analysis                        |
|                |                    |                  |                 | QG2-1061-1 |                                       |
|                |                    |                  |                 | QG2-1064-1 |                                       |
|                |                    |                  |                 | QG2-7180-1 |                                       |
|                |                    |                  |                 | QG2-7180-2 |                                       |
|                |                    |                  |                 | QG1-7246-1 |                                       |
